---
title: "Data Analysis with Quarto"
subtitle: "Descriptive statistics · plotting · linear regression · geostatistics"
lang: en
engine: knitr
execute:
  warning: false
  message: false
---

> 🌐 **Versi Bahasa Indonesia:** [baca dalam Bahasa Indonesia](../analisis.qmd)

This page shows real analysis — **run straight from the data** — in both **R** and
**Python**. Case study: poverty across 35 districts of Central Java
(`data/data_kemiskinan_jateng.csv`). Copy the code and run it on your machine.

```{r}
#| label: load-r
library(dplyr); library(ggplot2); library(knitr)
df <- read.csv("../data/data_kemiskinan_jateng.csv")
df23 <- df[df$tahun == 2023, ]
```

## 1. Descriptive statistics

Summaries of each variable: central tendency (mean, median) and spread (standard
deviation, range).

::: {.panel-tabset}
### R
```{r}
#| label: desc-r
vars <- c("kemiskinan", "pdrb_kapita", "inflasi", "ipm")
ringkas <- data.frame(
  Variable = vars,
  Mean = sapply(df23[vars], mean), Median = sapply(df23[vars], median),
  SD = sapply(df23[vars], sd),
  Min = sapply(df23[vars], min), Max = sapply(df23[vars], max))
kable(ringkas, digits = 2, row.names = FALSE)
```
### Python
```{python}
#| label: desc-py
import pandas as pd
df = pd.read_csv("../data/data_kemiskinan_jateng.csv")
df23 = df[df["tahun"] == 2023]
df23[["kemiskinan","pdrb_kapita","inflasi","ipm"]].describe().round(2)
```
:::

## 2. Plotting

Visualization reveals the **distribution** (histogram) and **relationships** (scatter).

::: {.panel-tabset}
### R
```{r}
#| label: plot-r
#| fig-cap: "Poverty distribution (histogram) & its relation to GDP per capita, 2023"
#| fig-width: 8
#| fig-height: 3.2
par(mfrow = c(1, 2))
hist(df23$kemiskinan, breaks = 10, col = "#b03a2e", border = "white",
     main = "", xlab = "Poverty (%)", ylab = "Count")
plot(df23$pdrb_kapita, df23$kemiskinan, pch = 19, col = "#2166ac",
     xlab = "GDP per capita", ylab = "Poverty (%)")
abline(lm(kemiskinan ~ pdrb_kapita, data = df23), col = "#b03a2e", lwd = 2)
```
### Python
```{python}
#| label: plot-py
#| fig-cap: "Poverty distribution (histogram) & its relation to GDP per capita, 2023"
#| fig-width: 8
#| fig-height: 3.2
import numpy as np, matplotlib.pyplot as plt
fig, ax = plt.subplots(1, 2, figsize=(8, 3.2))
ax[0].hist(df23["kemiskinan"], bins=10, color="#b03a2e", edgecolor="white")
ax[0].set_xlabel("Poverty (%)"); ax[0].set_ylabel("Count")
ax[1].scatter(df23["pdrb_kapita"], df23["kemiskinan"], color="#2166ac")
m, b = np.polyfit(df23["pdrb_kapita"], df23["kemiskinan"], 1)
ax[1].plot(df23["pdrb_kapita"], m*df23["pdrb_kapita"] + b, color="#b03a2e")
ax[1].set_xlabel("GDP per capita"); ax[1].set_ylabel("Poverty (%)")
plt.tight_layout(); plt.show()
```
:::

## 3. Multiple linear regression

Model poverty as a function of several predictors. Watch the **sign of each coefficient**
(direction), the **p-value** (significance), and **R²** (explained variance). We use the
**2019–2023 panel** (175 observations).

::: {.panel-tabset}
### R
```{r}
#| label: reg-r
model <- lm(kemiskinan ~ pdrb_kapita + inflasi + ipm + pengangguran, data = df)
koef <- summary(model)$coefficients[, c("Estimate", "Pr(>|t|)")]
rownames(koef) <- c("Intercept","GDP per capita","Inflation","HDI","Unemployment")
kable(koef, col.names = c("Coefficient", "p-value"), digits = 3)
cat(sprintf("R-squared = %.3f", summary(model)$r.squared))
```
### Python
```{python}
#| label: reg-py
import statsmodels.formula.api as smf
model = smf.ols("kemiskinan ~ pdrb_kapita + inflasi + ipm + pengangguran", data=df).fit()
koef = pd.DataFrame({"Coefficient": model.params, "p-value": model.pvalues}).round(3)
koef.index = ["Intercept","GDP per capita","Inflation","HDI","Unemployment"]
print(koef)
print(f"R-squared = {model.rsquared:.3f}")
```
:::

**Reading it:** a coefficient's **sign** is the direction (e.g. a negative HDI coefficient
⇒ higher HDI, lower poverty); **p-value < 0.05** ⇒ significant; **R²** = share of variance
explained. *A low R² for this simple model is expected* — panel data often needs **fixed
effects** or **spatial regression** (see the book). That's the point of being reproducible:
swap the model, re-render, compare.

## 4. Geostatistics

Spatial data carries an extra message: **do poor districts cluster?** We map the
distribution, then test for **spatial autocorrelation** with **Moran's I**.

### Choropleth map {.unnumbered}

::: {.panel-tabset}
### R
```{r}
#| label: map-r
#| fig-cap: "Poverty by district, 2023 (R)"
#| fig-width: 8
#| fig-height: 4.2
library(sf); library(maptiles); library(tidyterra)
geo <- st_transform(st_read("../data/jateng.geojson", quiet = TRUE), 3857)
peta <- merge(geo, df23, by = "wilayah", all.x = TRUE)
latar <- get_tiles(peta, provider = "OpenStreetMap", zoom = 8, crop = TRUE)
ggplot() + geom_spatraster_rgb(data = latar) +
  geom_sf(data = peta, aes(fill = kemiskinan), color = "white", linewidth = 0.1, alpha = 0.85) +
  scale_fill_distiller(palette = "OrRd", direction = 1, name = "Poverty (%)") + theme_void()
```
### Python
```{python}
#| label: map-py
#| fig-cap: "Poverty by district, 2023 (Python)"
#| fig-width: 8
#| fig-height: 4.2
import geopandas as gpd, contextily as ctx
g = gpd.read_file("../data/jateng.geojson")
pmap = g.merge(df23, on="wilayah", how="left").to_crs(3857)
ax = pmap.plot(column="kemiskinan", cmap="OrRd", legend=True, alpha=0.85,
               edgecolor="white", linewidth=0.3, figsize=(8, 4.2))
ctx.add_basemap(ax, source=ctx.providers.OpenStreetMap.Mapnik)
ax.set_axis_off(); plt.show()
```
:::

### Spatial autocorrelation — Moran's I {.unnumbered}

A positive, significant Moran's I ⇒ similar values sit **next to each other** (clustering).

::: {.panel-tabset}
### R
```{r}
#| label: moran-r
library(spdep)
nb <- poly2nb(peta)                 # neighbours (queen contiguity)
w  <- nb2listw(nb, style = "W")     # spatial weights
mt <- moran.test(peta$kemiskinan, w)
cat(sprintf("Moran's I = %.3f (p = %.4f)", mt$estimate[1], mt$p.value))
```
### Python
```{python}
#| label: moran-py
from libpysal.weights import Queen
from esda.moran import Moran
w = Queen.from_dataframe(pmap, use_index=False); w.transform = "r"
mi = Moran(pmap["kemiskinan"].values, w)
print(f"Moran's I = {mi.I:.3f} (p = {mi.p_sim:.4f})")
```
:::

**Interpretation:** if Moran's I is positive and significant, poverty **clusters
spatially** — there are poverty "pockets." Follow up with **LISA** (local clusters) and
**spatial regression** if needed (see the book's *Spatial Statistics* chapter).

## Next steps

- 📘 **Book "Quarto untuk Skripsi"** — classical assumption tests, panel data, LISA, spatial
  regression (Indonesian): <https://firmanhadi21.github.io/thesis-with-quarto/>
- 🧪 Back to the **[hands-on tutorial](hands-on.qmd)**.
