---
title: "Quarto for Scientific Writing"
subtitle: "Reproducible documents for students & researchers"
author: "Workshop · Zoom"
date: today
date-format: "D MMMM YYYY"
lang: en
engine: knitr
format:
  beamer:
    theme: Madrid
    aspectratio: 169
    fig-width: 6
    fig-height: 3.2
  revealjs:
    theme: simple
    slide-number: true
    fig-width: 7
    fig-height: 4
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr); library(ggplot2)
df <- read.csv("../data/data_kemiskinan_jateng.csv")
df23 <- df[df$tahun == 2023, ]
```

# Why Quarto?

## Who is this workshop for?

- 🎓 **Students** — quantitative thesis: data, tables, figures, citations, campus-formatted PDF.
- 🔬 **Lecturers & researchers** — **reproducible** papers, reports, and presentations.

\vspace{1em}
One tool, one workflow — from analysis to finished document.

## The old way hurts

| Word + SPSS/Excel | With Quarto |
|---|---|
| Copy numbers by hand → typos | Numbers **automatic** from code |
| Table/figure numbering breaks | Numbering & TOC **automatic** |
| Revise data = redo & re-paste | Change data → **re-render** |
| Manual citations | Automatic citations (APA, Zotero) |
| Slides built separately | Slides from the **same data** |

## What is Quarto?

A publishing system: **prose + code + data** in one `.qmd` file.

```
  .qmd  ──►  Quarto  ──►  PDF · HTML · Word · Slides
```

- Free & open source. Supports **R** and **Python**.
- **Write once, render many.**

# The core: data lives in the document

## Numbers & charts straight from data {.fragile}

The mean poverty rate in Central Java **2023 = `r round(mean(df23$kemiskinan),2)`%**
(computed at render time, not typed).

```{r}
#| fig-cap: "Mean poverty trend (from the data)"
df |> group_by(tahun) |> summarise(rata = mean(kemiskinan)) |>
  ggplot(aes(tahun, rata)) +
  geom_line(color = "#b03a2e", linewidth = 1) + geom_point(size = 2, color = "#b03a2e") +
  labs(x = "Year", y = "Poverty (%)") + theme_minimal(base_size = 13)
```

## Maps too, from the same data

```{r}
#| fig-cap: "Poverty by district, 2023"
#| fig-height: 3.0
library(sf); library(maptiles); library(tidyterra)
geo <- st_transform(st_read("../data/jateng.geojson", quiet = TRUE), 3857)
peta <- merge(geo, df23, by = "wilayah", all.x = TRUE)
latar <- get_tiles(peta, provider = "OpenStreetMap", zoom = 8, crop = TRUE)
ggplot() + geom_spatraster_rgb(data = latar) +
  geom_sf(data = peta, aes(fill = kemiskinan), color = "white", linewidth = 0.1, alpha = 0.85) +
  scale_fill_distiller(palette = "OrRd", direction = 1, name = "%") + theme_void()
```

## The killer feature: *reproducible* {.fragile}

> **Change one number in the data → the whole document updates:**
> inline numbers, tables, figures, **maps**, even **slides**.

\vspace{0.5em}
No more manual copy-paste. Advisor asks to change the year range? Re-render once.

## ⚠️ One gotcha

**`quarto render` — not just the *preview* button.**

- The live preview **does not watch data files**.
- After editing **data**, run **`quarto render`** and open the result.

# An analysis walkthrough

## Descriptive statistics

```{r}
vars <- c("kemiskinan", "pdrb_kapita", "inflasi", "ipm")
ringkas <- data.frame(Variable = vars,
  Mean = sapply(df23[vars], mean), Median = sapply(df23[vars], median),
  SD = sapply(df23[vars], sd))
knitr::kable(ringkas, digits = 2, row.names = FALSE)
```

Central tendency (mean, median) and spread (SD) for each variable — 2023.

## Plotting: relationships

```{r}
#| fig-cap: "Poverty vs GDP per capita (2023)"
#| fig-height: 2.9
plot(df23$pdrb_kapita, df23$kemiskinan, pch = 19, col = "#2166ac",
     xlab = "GDP per capita", ylab = "Poverty (%)")
abline(lm(kemiskinan ~ pdrb_kapita, data = df23), col = "#b03a2e", lwd = 2)
```

## Multiple linear regression

```{r}
model <- lm(kemiskinan ~ pdrb_kapita + inflasi + ipm + pengangguran, data = df)
koef <- summary(model)$coefficients[, c("Estimate", "Pr(>|t|)")]
rownames(koef) <- c("Intercept","GDP/capita","Inflation","HDI","Unemployment")
knitr::kable(koef, col.names = c("Coefficient", "p-value"), digits = 3)
```

\footnotesize **R² = `r round(summary(model)$r.squared, 3)`**. Read the coefficient
**sign** (direction) & **p-value** (significance). Low R² $\Rightarrow$ needs fixed effects / spatial
regression.

## Geostatistics: spatial autocorrelation

Does poverty **cluster** spatially? Test with **Moran's I**:

```{r}
library(spdep)
nb <- poly2nb(peta); w <- nb2listw(nb, style = "W")
mt <- moran.test(peta$kemiskinan, w)
```

**Moran's I = `r round(mt$estimate[1], 3)`** (p = `r round(mt$p.value, 4)`).
Positive & significant $\Rightarrow$ poverty **"pockets"** (spatial clusters). Next: LISA & spatial
regression.

# For your work

## Students: a pre-formatted thesis

- **Chapter I–V** template, cover page, approval page — to campus guidelines.
- "BAB I" numbering, **Table 4.1 / Figure 4.2**, TOC/lists/bibliography — automatic.
- Write the content, swap in your data, hit **Render** → thesis PDF.

## Researchers: reproducible writing & talks

- **Papers & reports**: one source → PDF/Word/HTML; citations via **Zotero**.
- **Presentations**: **Reveal.js** (interactive HTML) & **Beamer** (PDF) from one file.
- Collaborate via **Git**; publish to **GitHub Pages** automatically.

## One source, two kinds of slides

```yaml
format:
  revealjs: default     # HTML slides (screen/web)
  beamer:               # PDF slides (projector)
    aspectratio: 169
```

```bash
quarto render slide.qmd --to revealjs   # -> slide.html
quarto render slide.qmd --to beamer     # -> slide.pdf
```

*These very slides are made that way — Beamer for PDF, Reveal.js for the web.*

# Get started now

## How to start

1. Install **R + RStudio** (Quarto is included).
2. `quarto install tinytex` (for PDF).
3. *(Optional)* Python + `reticulate` for Python examples.
4. Create `hello.qmd`, hit **Render**. Done.

## Hands-on & resources

- 🧪 **Hands-on tutorial (1 hour):** this workshop site (*Hands-on* menu).
- 📘 **Book "Quarto untuk Skripsi" (Indonesian):** \newline <https://firmanhadi21.github.io/thesis-with-quarto/>
- 📑 **Quarto docs:** <https://quarto.org>

## {.center}

\begin{center}
{\Huge Thank You}\\[0.5em]
{\large Let's write reproducibly.}
\end{center}
